---
title: "Commodity Analysis"
author: "Caleb King"
date: "`r format(Sys.time(),  '%B %d, %Y')`"
output:
  pdf_document: default
fontsize: 12pt
geometry: margin=1.25cm
header-includes:
- \usepackage{amssymb}
- \usepackage{dsfont}
- \let\rmarkdownfootnote\footnote%
- \def\footnote{\protect\rmarkdownfootnote}
- \usepackage{titling}
- "\\setlength{\\droptitle}{-2em}"
- \pretitle{\vspace{\droptitle}\centering\huge}
- \posttitle{\par}
- \preauthor{\centering\large\emph}
- \postauthor{\par}
- \predate{\centering\large\emph}
- \postdate{\par}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, warning=FALSE}
# Load necessary libraries
library(tseries)
library(forecast)
library(urca)
library(CommoditiesPackage)
```

Get data for commodities needed

Monthly Precious Metal
```{r}
head(commodityData$iPRECIOUSMET)
```
Translate to time series format

Monthly Precious metal
```{r}
# Create a time series of Precious metal monthly level
metalMonth.t <- ts(commodityData$iPRECIOUSMET,start=1960,freq=12 )   
#transform data
metalMonth.t <- log(metalMonth.t)
tail(metalMonth.t) 
```
Numerical summary

Sample size
```{r}
# Sample size
n <- length( metalMonth.t )
n
```


Mean and standard deviation
```{r}
mean(metalMonth.t)
sd(metalMonth.t)
```


Five-number (minimum, 1st quartile, median, 3rd quartile, maximum) summary
```{r}
# Five-number (minimum, 1st quartile, median, 3rd quartile, maximum) summary and mean
summary( metalMonth.t )
```


Visualization
We use a time plot to visualize a time series

Nominal US dollar term
```{r}
# Time plot using plot.ts()
plot.ts(metalMonth.t, ylim=c(1,6), col="#00AFBB", xlab="Year", ylab="Price", main="Precious Metal Monthly Price log(index)")
legend("bottomright", legend="Source: World Bank", bty="n", cex=0.8, col="gray40")
```


Test for stationary or not
```{r}
adf.test(metalMonth.t, k=12)
```


Sample ACF and PACF of time series data
```{r}
# Sample ACF and PACF
par( mfrow=c(2,1),mex=0.75 )
acf( metalMonth.t,lag.max=144,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( metalMonth.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Taking the differencing for seasonality because it is monthly data
```{r}
# Lag-12 differenced series (Note: the sample size is decreased by 12 due to lag-12 differencing)
x.t_s12 <- diff(metalMonth.t, 12)

# Time plot of lag-12 difference series
plot.ts( x.t_s12,ylim=c(-2,2),xlab="Year",main="Lag-12 differenced Metal Data" )

# Sample ACF and PACF of lag-12 differenced series
par( mfrow=c(2,1),mex=0.75 )
acf( x.t_s12,lag.max=144,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( x.t_s12,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Numerical Summary of Lag-12 differenced data
```{r}
# Numerical summary of lag-12 differenced series
x.t_s12.mean <- mean( x.t_s12 )
x.t_s12.sd <- sd( x.t_s12 )
n_s12 <- length( x.t_s12 )

x.t_s12.mean
x.t_s12.sd
n_s12

summary( x.t_s12 )
```


Test for Linear trend on differenced data
```{r}
# Augmented Dickey-Fuller test for unit roots
adf.out_0 <- ur.df( x.t_s12,type="drift",lags=0 )         # Augmented Dickey-Fuller test with p=0 ("lags=0")
summary( adf.out_0 )
```
Because the ADF test statistic value -4.284 is less than the 5th percentile critical value -2.86, at  alpha=0.05  significance level, we do reject the null hypothesis  H0 : the series x.t has a unit root. Therefore, it is not supported to apply the lag-1 differencing to the lag-12 differenced series metalMonth.t.


Function to compute AICC
```{r}
# Function to compute AICC
arma.AICC <- function(fit) {
  n.par <- length(fit$coef)-sum((fit$coef==0))     # no of parameters in ARMA fit, corrected for sub-models
  n <- fit$nobs                                    # sample size
  aicc <- -2*fit$loglik+2*(n.par+1)*n/(n-n.par-2)  # AICC
  return(aicc)
}
```


Fit an appropriate model
```{r}
fit_101.210 <- arima(metalMonth.t,order=c(1,0,1),seasonal=list(order=c(2,1,0), period=12))
fit_101.210
```


Analyze the BIC and AICC values (smaller values are better)
```{r}
arma.AICC(fit_101.210)
BIC(fit_101.210)
```


Create variable for the best fitting data
```{r}
fit.bst <- fit_101.210
```


Compute residuals and summary of residuals
```{r}
# Residuals
w.t <- ts( fit.bst$resid[13:n],start=c(1960,1),freq=12 )
head( w.t )

# Numerical summary of residuals
w.mean <- mean(w.t)                              # should be close to zero
w.sd <- sd(w.t)                                  # should be close to sigma

w.mean
w.sd

summary(w.t)
```


Model diagnostics 
```{r}
# Model diagnostics: visual inspection
par( mfrow=c(3,1),mex=0.75 )
plot.ts( w.t,ylim=c(min(w.t),max(w.t)),xlab="Year",ylab="Residuals",main="Model Diagnostics" )
abline( h=0,col="blue" )
acf( w.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( w.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Ljung-Box test
```{r}
# Model diagnostics: Ljung-Box portmanteau test (repeat with different m)
m <- 10                                                           # Ljung-Box portmanteau test's "m"
n.par <- length(fit.bst$coef)                                     # number of parameters in fitted ARMA(p,q), p+q
LB.out <- Box.test(w.t,lag=m,type="Ljung-Box")                    # Ljung-Box portmanteau test
LB.out                                                            # Box.test uses wrong df if applied to residuals
LB.pval <- pchisq(q=LB.out$statistic,df=m-n.par,lower.tail=FALSE) # correct p-value
round( LB.pval, 4 )
```


Normality Check
```{r}
# Model diagnostics: normality check
par(mfrow=c(2,1),mex=0.75)
hist(w.t,freq=FALSE,                                              # histogram of residuals
     breaks=seq(-1,1,0.05),
     col="grey85",ylim=c(0,10),
     main="Residual Histogram")
z <- seq(-8,8,length=1000)
lines(z,dnorm(z,mean=w.mean,sd=w.sd),lty=1,col="red")             # add theoretical normal density
qqnorm(w.t)                                                       # normal Q-Q plot
qqline(w.t)
```


Normality test (Shapiro-Wilk and Kolmogorov-Smirnov tests)
```{r}
# Normality test
shapiro.test(w.t)                                                 # Shapiro-Wilk normality test
ks.test(w.t,"pnorm",mean=w.mean,sd=w.sd)                          # Kolmogorov-Smirnov normality test
```


Furure value prediction (Forecast)
```{r}
# Forecast 12 future values of original series x.t with 95% PI (only when model diagnostics is passed)
x.fct <- predict( fit.bst,n.ahead=14 )
x.L95 <- x.fct$pred-1.96*x.fct$se
x.U95 <- x.fct$pred+1.96*x.fct$se

par( mfrow=c(1,1), mex=0.75 )
ts.plot( metalMonth.t,x.fct$pred, col=c("black","tomato"),ylim=c(0,6),xlim=c(2001,2028),
         main="Monthly Precious Metal Price with 14 Months Predicted" )
lines( x.U95,col="cyan4",lty="dashed" )
lines( x.L95,col="cyan4",lty="dashed" )
legend( "bottom", lty=c("solid","solid","dashed"),
        legend=c("Actual", "Predicted", "Upper and Lower 95%"), col=c("black", "tomato", "cyan4"))
```

Monthly Energy
```{r}
head(commodityData$iENERGY)
```


Translate to time series format

Monthly Energy
```{r}
# Create a time series of Precious energy monthly level
energyMonth.t <- ts(commodityData$iENERGY,start=1960,freq=12 )   
#transform data
energyMonth.t <- log(energyMonth.t)
tail(energyMonth.t) 
```


Numerical summary

Sample size
```{r}
# Sample size
n <- length( energyMonth.t )
n
```


Mean and standard deviation
```{r}
mean(energyMonth.t)
sd(energyMonth.t)
```


Five-number (minimum, 1st quartile, median, 3rd quartile, maximum) summary
```{r}
# Five-number (minimum, 1st quartile, median, 3rd quartile, maximum) summary and mean
summary( energyMonth.t )
```


Visualization
We use a time plot to visualize a time series

Nominal US dollar term
```{r}
# Time plot using plot.ts()
plot.ts(energyMonth.t, ylim=c(0,6), col="#00AFBB", xlab="Year", ylab="Price", main="Precious Energy Monthly Price log(index)")
legend("bottomright", legend="Source: World Bank", bty="n", cex=0.8, col="gray40")
```


Test for stationary or not
```{r}
adf.test(energyMonth.t, k=12)
```


Sample ACF and PACF of time series data
```{r}
# Sample ACF and PACF
par( mfrow=c(2,1),mex=0.75 )
acf( energyMonth.t,lag.max=144,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( energyMonth.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Taking the differencing for seasonality because it is monthly data
```{r}
# Lag-12 differenced series (Note: the sample size is decreased by 12 due to lag-12 differencing)
x.t_s12 <- diff(energyMonth.t, 12)

# Time plot of lag-12 difference series
plot.ts( x.t_s12,ylim=c(-2,2),xlab="Year",main="Lag-12 differenced Energy Data" )

# Sample ACF and PACF of lag-12 differenced series
par( mfrow=c(2,1),mex=0.75 )
acf( x.t_s12,lag.max=144,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( x.t_s12,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Numerical Summary of Lag-12 differenced data
```{r}
# Numerical summary of lag-12 differenced series
x.t_s12.mean <- mean( x.t_s12 )
x.t_s12.sd <- sd( x.t_s12 )
n_s12 <- length( x.t_s12 )

x.t_s12.mean
x.t_s12.sd
n_s12

summary( x.t_s12 )
```


Test for Linear trend on differenced data
```{r}
# Augmented Dickey-Fuller test for unit roots
adf.out_0 <- ur.df( x.t_s12,type="drift",lags=0 )         # Augmented Dickey-Fuller test with p=0 ("lags=0")
summary( adf.out_0 )
```
Because the ADF test statistic value -4.5813 is less than the 5th percentile critical value -2.86, at  alpha=0.05  significance level, we do reject the null hypothesis  H0 : the series x.t has a unit root. Therefore, it is not supported to apply the lag-1 differencing to the lag-12 differenced series energyMonth.t.


Function to compute AICC
```{r}
# Function to compute AICC
arma.AICC <- function(fit) {
  n.par <- length(fit$coef)-sum((fit$coef==0))     # no of parameters in ARMA fit, corrected for sub-models
  n <- fit$nobs                                    # sample size
  aicc <- -2*fit$loglik+2*(n.par+1)*n/(n-n.par-2)  # AICC
  return(aicc)
}
```


Fit an appropriate model
```{r}
fit_200.210 <- arima(energyMonth.t,order=c(2,0,0),seasonal=list(order=c(2,1,0), period=12))
fit_200.210
```


Analyze the BIC and AICC values (smaller values are better)
```{r}
arma.AICC(fit_200.210)
BIC(fit_200.210)
```


Create variable for the best fitting data
```{r}
fit.bst <- fit_200.210
```


Compute residuals and summary of residuals
```{r}
# Residuals
w.t <- ts( fit.bst$resid[13:n],start=c(1960,1),freq=12 )
head( w.t )

# Numerical summary of residuals
w.mean <- mean(w.t)                              # should be close to zero
w.sd <- sd(w.t)                                  # should be close to sigma

w.mean
w.sd

summary(w.t)
```


Model diagnostics 
```{r}
# Model diagnostics: visual inspection
par( mfrow=c(3,1),mex=0.75 )
plot.ts( w.t,ylim=c(min(w.t),max(w.t)),xlab="Year",ylab="Residuals",main="Model Diagnostics" )
abline( h=0,col="blue" )
acf( w.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( w.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Ljung-Box test
```{r}
# Model diagnostics: Ljung-Box portmanteau test (repeat with different m)
m <- 10                                                           # Ljung-Box portmanteau test's "m"
n.par <- length(fit.bst$coef)                                     # number of parameters in fitted ARMA(p,q), p+q
LB.out <- Box.test(w.t,lag=m,type="Ljung-Box")                    # Ljung-Box portmanteau test
LB.out                                                            # Box.test uses wrong df if applied to residuals
LB.pval <- pchisq(q=LB.out$statistic,df=m-n.par,lower.tail=FALSE) # correct p-value
round( LB.pval, 4 )
```


Normality Check
```{r}
# Model diagnostics: normality check
par(mfrow=c(2,1),mex=0.75)
hist(w.t,freq=FALSE,                                              # histogram of residuals
     breaks=seq(-1,1,0.05),
     col="grey85",ylim=c(0,10),
     main="Residual Histogram")
z <- seq(-8,8,length=1000)
lines(z,dnorm(z,mean=w.mean,sd=w.sd),lty=1,col="red")             # add theoretical normal density
qqnorm(w.t)                                                       # normal Q-Q plot
qqline(w.t)
```


Normality test (Shapiro-Wilk and Kolmogorov-Smirnov tests)
```{r}
# Normality test
shapiro.test(w.t)                                                 # Shapiro-Wilk normality test
ks.test(w.t,"pnorm",mean=w.mean,sd=w.sd)                          # Kolmogorov-Smirnov normality test
```


Furure value prediction (Forecast)
```{r}
# Forecast 12 future values of original series x.t with 95% PI (only when model diagnostics is passed)
x.fct <- predict( fit.bst,n.ahead=14 )
x.L95 <- x.fct$pred-1.96*x.fct$se
x.U95 <- x.fct$pred+1.96*x.fct$se

par( mfrow=c(1,1), mex=0.75 )
ts.plot( energyMonth.t,x.fct$pred, col=c("black","tomato"),ylim=c(0,6),xlim=c(2001,2026),
         main="Monthly Energy Price with 14 Months Predicted" )
lines( x.U95,col="cyan4",lty="dashed" )
lines( x.L95,col="cyan4",lty="dashed" )
legend( "bottom", lty=c("solid","solid","dashed"),
        legend=c("Actual", "Predicted", "Upper and Lower 95%"), col=c("black", "tomato", "cyan4"))
```


Monthly Precious nonfuel
```{r}
head(commodityData$iNONFUEL)
```


Translate to time series format

Monthly Precious nonfuel
```{r}
# Create a time series of Precious metal monthly level
nonfuelMonth.t <- ts(commodityData$iNONFUEL,start=1960,freq=12 )   
#transform data
nonfuelMonth.t <- log(nonfuelMonth.t)
tail(nonfuelMonth.t) 
```


Numerical summary

Sample size
```{r}
# Sample size
n <- length( nonfuelMonth.t )
n
```


Mean and standard deviation
```{r}
mean(nonfuelMonth.t)
sd(nonfuelMonth.t)
```


Five-number (minimum, 1st quartile, median, 3rd quartile, maximum) summary
```{r}
# Five-number (minimum, 1st quartile, median, 3rd quartile, maximum) summary and mean
summary( nonfuelMonth.t )
```


Visualization
We use a time plot to visualize a time series

Nominal US dollar term
```{r}
# Time plot using plot.ts()
plot.ts(nonfuelMonth.t, ylim=c(1,6), col="#00AFBB", xlab="Year", ylab="Price", main="Precious nonfuel Monthly Price log(index)")
legend("bottomright", legend="Source: World Bank", bty="n", cex=0.8, col="gray40")
```


Test for stationary or not
```{r}
adf.test(nonfuelMonth.t, k=12)
```


Sample ACF and PACF of time series data
```{r}
# Sample ACF and PACF
par( mfrow=c(2,1),mex=0.75 )
acf( nonfuelMonth.t,lag.max=144,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( nonfuelMonth.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Taking the differencing for seasonality because it is monthly data
```{r}
# Lag-12 differenced series (Note: the sample size is decreased by 12 due to lag-12 differencing)
x.t_s12 <- diff(nonfuelMonth.t, 12)

# Time plot of lag-12 difference series
plot.ts( x.t_s12,ylim=c(-2,2),xlab="Year",main="Lag-12 differenced Non-Energy Data" )

# Sample ACF and PACF of lag-12 differenced series
par( mfrow=c(2,1),mex=0.75 )
acf( x.t_s12,lag.max=144,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( x.t_s12,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Numerical Summary of Lag-12 differenced data
```{r}
# Numerical summary of lag-12 differenced series
x.t_s12.mean <- mean( x.t_s12 )
x.t_s12.sd <- sd( x.t_s12 )
n_s12 <- length( x.t_s12 )

x.t_s12.mean
x.t_s12.sd
n_s12

summary( x.t_s12 )
```


Test for Linear trend on differenced data
```{r}
# Augmented Dickey-Fuller test for unit roots
adf.out_0 <- ur.df( x.t_s12,type="drift",lags=0 )         # Augmented Dickey-Fuller test with p=0 ("lags=0")
summary( adf.out_0 )
```
Because the ADF test statistic value -4.0586 is less than the 5th percentile critical value -2.86, at  alpha=0.05  significance level, we do reject the null hypothesis  H0 : the series x.t has a unit root. Therefore, it is not supported to apply the lag-1 differencing to the lag-12 differenced series nonfuelMonth.t.


Function to compute AICC
```{r}
# Function to compute AICC
arma.AICC <- function(fit) {
  n.par <- length(fit$coef)-sum((fit$coef==0))     # no of parameters in ARMA fit, corrected for sub-models
  n <- fit$nobs                                    # sample size
  aicc <- -2*fit$loglik+2*(n.par+1)*n/(n-n.par-2)  # AICC
  return(aicc)
}
```


Fit an appropriate model
```{r}
fit_101.210 <- arima(nonfuelMonth.t,order=c(2,0,1),seasonal=list(order=c(2,1,0), period=12))
fit_101.210
```


Analyze the BIC and AICC values (smaller values are better)
```{r}
arma.AICC(fit_101.210)
BIC(fit_101.210)
```


Create variable for the best fitting data
```{r}
fit.bst <- fit_101.210
```


Compute residuals and summary of residuals
```{r}
# Residuals
w.t <- ts( fit.bst$resid[13:n],start=c(1960,1),freq=12 )
head( w.t )

# Numerical summary of residuals
w.mean <- mean(w.t)                              # should be close to zero
w.sd <- sd(w.t)                                  # should be close to sigma

w.mean
w.sd

summary(w.t)
```


Model diagnostics 
```{r}
# Model diagnostics: visual inspection
par( mfrow=c(3,1),mex=0.75 )
plot.ts( w.t,ylim=c(min(w.t),max(w.t)),xlab="Year",ylab="Residuals",main="Model Diagnostics" )
abline( h=0,col="blue" )
acf( w.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample ACF",main="" )
pacf( w.t,lag.max=24,ylim=c(-0.5,1),ylab="Sample PACF",main="" )
```


Ljung-Box test
```{r}
# Model diagnostics: Ljung-Box portmanteau test (repeat with different m)
m <- 10                                                           # Ljung-Box portmanteau test's "m"
n.par <- length(fit.bst$coef)                                     # number of parameters in fitted ARMA(p,q), p+q
LB.out <- Box.test(w.t,lag=m,type="Ljung-Box")                    # Ljung-Box portmanteau test
LB.out                                                            # Box.test uses wrong df if applied to residuals
LB.pval <- pchisq(q=LB.out$statistic,df=m-n.par,lower.tail=FALSE) # correct p-value
round( LB.pval, 4 )
```


Normality Check
```{r}
# Model diagnostics: normality check
par(mfrow=c(2,1),mex=0.75)
hist(w.t,freq=FALSE,                                              # histogram of residuals
     breaks=seq(-1,1,0.05),
     col="grey85",ylim=c(0,10),
     main="Residual Histogram")
z <- seq(-8,8,length=1000)
lines(z,dnorm(z,mean=w.mean,sd=w.sd),lty=1,col="red")             # add theoretical normal density
qqnorm(w.t)                                                       # normal Q-Q plot
qqline(w.t)
```


Normality test (Shapiro-Wilk and Kolmogorov-Smirnov tests)
```{r}
# Normality test
shapiro.test(w.t)                                                 # Shapiro-Wilk normality test
ks.test(w.t,"pnorm",mean=w.mean,sd=w.sd)                          # Kolmogorov-Smirnov normality test
```


Furure value prediction (Forecast)
```{r}
# Forecast 12 future values of original series x.t with 95% PI (only when model diagnostics is passed)
x.fct <- predict( fit.bst,n.ahead=14 )
x.L95 <- x.fct$pred-1.96*x.fct$se
x.U95 <- x.fct$pred+1.96*x.fct$se

par( mfrow=c(1,1), mex=0.75 )
ts.plot( nonfuelMonth.t,x.fct$pred, col=c("black","tomato"),ylim=c(0,6),xlim=c(2001,2028),
         main="Monthly nonfuel Price with 14 Months Predicted" )
lines( x.U95,col="cyan4",lty="dashed" )
lines( x.L95,col="cyan4",lty="dashed" )
legend( "bottom", lty=c("solid","solid","dashed"),
        legend=c("Actual", "Predicted", "Upper and Lower 95%"), col=c("black", "tomato", "cyan4"))
```


